Gaps in stellar streams as a result of globular cluster fly-bys

### 2024-11-19 paper/main.tex
swapped results and conclusions for Paola's rewrite

  \subsection{Gap Detection and Characterization} \label{sec:gap_methods}

    At the end of the simulations, some gaps in the stellar stream are so prominent that they can be identified visually without any sophisticated quantitative analysis. The analysis of the morphology of the tidal tails of Palomar~5 is carried out by comparing simulations with globular clusters (called \texttt{full}) and without them (called \texttt{reference}). A preliminary evaluation is performed visually, followed by a quantitative analysis of the differences between the density profiles in the two mentioned cases.

    The \texttt{reference} simulations used the same 50 sets of initial conditions but excluded interactions with globular clusters, except for the host cluster, Palomar 5. In other words, the gravitational influence of all other clusters on the star particles within the stream was omitted. By comparing the morphological differences between these \texttt{reference} simulations and the \texttt{full} simulations, we could more effectively identify and assess the impact of globular clusters on the stream's structure. Details on gap identification are provided in Appendix~\ref{sec:gap_detection}, and the identification of corresponding perturbing globular clusters is detailed in Appendix~\ref{sec:Perturber_Identification}.

    Comparisons between the \texttt{reference} and \texttt{full} simulations were performed in the tail coordinate system. This system allows for consistent morphological comparisons by aligning the stream with the cluster's orbit. The coordinate system is based on the work of \citet{2004AJ....127.2753D} and is shown in Fig.~\ref{fig:TailCoordinates}. Briefly, the x' coordinate represents the position of a particle along the orbit relative to the globular cluster—whether it is ahead or behind the cluster. The y' coordinate measures the particle's distance within the orbital plane, where positive values indicate that the particle is farther from the Galactic center and negative values indicate that it is closer.

    Once the gaps and perturbers were identified, we sought to characterize the properties of the globular clusters that caused gaps and differentiate them from those that did not. \citet{2013ApJ...775...90C} initiated the theory of gap creation in stellar streams due to gravitational flybys. This analysis was extended both analytically and numerically by \citet{2015MNRAS.450.1136E}, who studied uniform-density streams on circular orbits. \citet{2016MNRAS.457.3817S} presented the most comprehensive theory, generalizing Erkal et al.'s work to include impacts on streams of varying densities on inclined and eccentric orbits using an action-angle approach.

    The significance of an impact is measured by the change in momentum imparted to the stream stars by the perturber. A useful approximation for the change in momentum is the force at the closest approach multiplied by an estimate of the interaction time: \begin{equation} \label{eq:change_in_momentum} \Delta v \approx \text{Force} \times \text{interaction time} = \frac{GM}{b^2} \times 2\frac{b}{v} = 2\frac{GM}{bv}, \end{equation} where $M$ is the mass of the perturber, $b$ is the impact parameter, $v$ is the relative velocity of the perturber with respect to the particle, and $G$ is the gravitational constant. We compute this for the five closest approaches of all globular clusters throughout each of the 50 simulations. Details on how this computation was performed are given in Appendix~\ref{sec:reconstruction}. Ultimately, we obtained a sample of 250 impacts and compared the impact geometry of those that caused gaps with those that did not.

  \subsection*{Overview}

@@

    To our knowledge, this is the first study to investigate interactions between globular clusters in this context. We included N-body forces on the orbits with the globular clusters, yet did not observe significant changes to orbital properties, such as energy or angular momentum, between any of the globular clusters in any simulation. This result is expected given the small sizes and vast distances between the clusters.

    We inspected all simulations upon their completion and present one particularly clear example shown in Fig.~\ref{fig:stream_on_sky} which we selected for its prominent gaps. Two of these gaps are visible in galactic coordinates and become even more apparent when marginalizing over latitude to view their 1D density profile over longitude.

    We applied the gap detection methods described in Appendix~\ref{sec:gap_detection}, along with the perturber identification methods from Appendix~\ref{sec:Perturber_Identification}. Results from all simulations are presented in Appendix~\ref{sec:gallery_of_gaps}. Once we identified the perturbers responsible for each gap, we conducted additional simulations to verify correct identification of the perturbers. These additional contain only the suspected clusters and were launched one-at-a-time to isolate the effects of a specific clusters. These results are shown in Fig.~\ref{fig:decomposition}. A summary of which perturber caused a gap and when is presented in Fig.~\ref{fig:histogram_impact_time}.

@@

    NGC~2808 impacts Palomar 5's stream about 200 Myr ago in 44 out of 50 simulations, creating a small gap at a similar position in each case. The gap size aligns with gap formation as a dispersion process, where stars gradually spread apart due to slight differences in orbital frequencies. Smaller gaps indicate more recent encounters. Since this interaction occurred less than one orbital period ago, the orbital solutions remain similar, producing consistent results across simulations.

    Most gaps appear in the leading tail, which we consider coincidental, as stars escape the cluster's L1 and L2 points at similar rates, and the small distance difference between tails' galactocentric distance is negligible. We expect gaps to be equally likely in both tails. However, only 8 of the 73 gaps occur in the trailing tail, with 44 linked to NGC2808's flyby. After removing these, there are 21 gaps in the leading tail and 8 in the trailing. By the binomial distribution, the probability of observing up to 8 successes in 29 trials, given a 50\% chance of success, is 1.2\%—unlikely, but possible. Additionally, other perturbers impact at consistent times, which may also violate the assumption of independent events, as seen with NGC2808.

    Overall, across 50 simulations, we observe 73 gaps. The distribution of simulations with a given number of gaps is provided in Table~\ref{table:gap_distribution}.  If we consider NGC 2808 to be an outlier and exclude it from the experiment, we observe an average of 0.6 gaps per simulation. Including all observed gaps results in an average of 1.5 gaps per simulation.

    \begin{table}[h]
      \centering
        \begin{tabular}{|l|c|c|c|c|c|}
          \hline
          Number of Gaps & 0 & 1 & 2 & 3 & 4 \\
          \hline
          Number of Sims. & 4 & 25 & 16 & 4 & 1 \\
          \hline
        \end{tabular}
      \caption{The gap occurrences.}\label{table:gap_distribution}
    \end{table}

 \subsection*{Perturber Statistics}

    We aim to identify general characteristics of perturbers that distinguish them from globular clusters that do not affect the Palomar 5 stream. Fig.~\ref{fig:mass_size_plane} presents several analyses to explore this.

    \begin{figure*}
      \centering
      \includegraphics[width=0.45\linewidth]{E_Lz_perturbers.png}
      \includegraphics[width=0.45\linewidth]{mass_size_plane.png}

      \includegraphics[width=0.45\linewidth]{impact_geometry_statistics_deltaP.png}
      \includegraphics[width=0.45\linewidth]{impact_stats_phase_space.png}
      \caption{\textbf{Characterisitcs of gap causing clusters}--\textit{Note} that the colors of the markers and histogram bars are consistent with Fig.~\ref{fig:histogram_impact_time}.\\ \textbf{Top Left:} Energy-angular momentum space of the globular clusters in the simulations, with 50~$\times$~165 data points representing all sampled initial conditions. Clusters impacting Palomar 5 are shown with colored markers; non-gap-causing clusters are gray. Palomar 5 is marked with a white symbol and black edge. The circular velocity-curve is drawn with a light-pink dashed curve.\\ \textbf{Top Right:} The mass-size plane. Uncertainties on the mass are indicated with vertical bars. \\ \textbf{Bottom Left:} The distribution of imparted momentum from a cluster flyby given by Eq.~\ref{eq:change_in_momentum}. The data set includes the top 5 strongest flybys from each simulation. Those that cause gaps are colored while those that do not are left in black. \\ \textbf{Bottom Right:} Palomar 5's orbit in the meridional plane. The grayscale represents all 50 stacked orbits, with $\mathcal{P}$ indicating the probability of  Palomar~5's position, normalized to $\mathcal{P}_\textrm{max}=1$. Markers show the perturber's position during each impact.}
      \label{fig:mass_size_plane}
      \end{figure*}

    The top-left panel shows that all clusters impacting the stream fall within a specific range of orbital energies, between $-2$ and $-1 \times10^5~\textrm{km}^2\textrm{s}^{-2}$. This is expected, as only clusters within Palomar 5's orbital space can interact with the stream. Our results are consistent with the findings from \citet{2021MNRAS.505.5978V}, who identified a wide range of orbital solutions for the least bound clusters, including both retrograde and prograde orbits. In our potential model, only three clusters samplings had positive orbital energies.

    The mass-size relation, shown in the top-right panel, reveals a lower limit of $10^5~\textrm{M}_{\odot}$ for clusters that provoke gaps. While mass and size alone are insufficient predictors, this empirical limit is useful because relative velocity and impact parameters are inacessible. Concentration or the density profile of the perturber has a much less significant effect respect to the total mass and subsequent evolution of the gap as shown in Fig.~3 of \citet{2016MNRAS.457.3817S}. The bottom-left panel offers a more physically meaningful result: gap-creating impacts tend to impart more momentum into the stream than non-gap impacts, though with some overlap. While \citet{2015MNRAS.450.1136E} extended the three-parameter model (Eq.~\ref{eq:change_in_momentum}) to include the size of the perturber and the perpendicular and parallel components of relative velocity, we found no clear combination of these factors that distinguishes gap from non-gap cases.

    Further insights are shown in Fig.\ref{fig:impact_geometry_statistics}, highlighting a maximum impact parameter of around 200~pc, roughly the width of the stream. Additionally, as \citet{2016MNRAS.457.3817S} demonstrated, the dispersion of action-frequencies in the stream plays a role. The frequencies being the rate at which a particle completes a cycle in phase space. For instance, the frequency corresponding to the azimuthal action, $J_\phi$, is $\dot\theta_\phi = - \frac{\mathcal{\partial H}}{\partial J_\phi}$ and where $\theta_\phi$ is the angle describing the position of the particle phase space between momentum ($p_\phi$) and position ($\phi$) axes---where $\phi$ is the azimuhtal angle between the x-y axes in physical space. Regions of the stream with well-separated frequencies are more susceptible to gap formation, while those with a wide frequency range (near the cluster) tend to erase the history of impacts. Thus, for a gap to form, the imparted change in frequency must exceed the range of frequencies in the impacted region. Thus, the strong flybys that occured close to the cluster and were thus inconsequential for gap formation.

    The bottom-right panel reveals a strong negative correlation between the location of impacts and the galactocentric radius, $r$. The number of gap creating impacts follows the relation $N = -2.5r + 50$, with a Pearson coefficient of -0.86. This is a non-obvious result. While there are more clusters near the galactic center, orbits naturally spend more time near their apocenters, and the lower relative velocities in these regions should favor gap creation. However, this result suggests that cluster galactic number density outweighs these factors in determining the number of gaps.

    Lastly, \citet{2012ApJ...748...20C} conceptualized the number of gaps using a statistical quantity: the rate at which gaps are created per unit stream length. For our simulations, this is given by: \begin{equation} \label{eq:gap_creation_rate}
    \mathcal{R}_{\textrm{Pal 5}} =  \frac{1}{T}\int_{0}^T l^{-1}(t) \sum_i \delta(t-t_i) dt,\end{equation}where $T$ is the total integration time, $l(t)$ is the length of the stream, and $\sum_i \delta(t-t_i)$ sums over the gap occurrences, with $i$ indexing over the number of gaps in a given simulation. Here, $\delta$ represents the diract delta function. This expression can be simplified to:\begin{equation}\mathcal{R}_{\textrm{Pal 5}} =  \frac{1}{T} \sum_i \frac{1}{l (t_i)}. \end{equation}
    This computation allows us to analyze the distribution of gap creation rates across all simulations. The results are shown in Fig.~\ref{fig:TailCoordinates}. This is roughly consistent with a simple estimate of the average gap creation rate: with 72 gaps over 5 Gyr of integration time for a stream about 20 kpc in length, the naive estimate is approximately 0.014~[km~s$^{-1}$~kpc$^{-2}$]. This is about double the weighted mean gap creation rate of 0.009~[km~s$^{-1}$~kpc$^{-2}$]. The naive estimate is higher because it does not account for the growth of the stream over time, unlike Eq.\ref{eq:gap_creation_rate}, where the likelihood of impacting a shorter stream is lower.

      \begin{figure}
        \centering
        \includegraphics[width=\linewidth]{gap_creation_rate.png}
        \caption{The distribution of the number of gaps normalized over the total integration time and unit stream length, as described by Eq.\ref{eq:gap_creation_rate}. }
        \label{fig:TailCoordinates}
      \end{figure}

>>>>

  \subsection{Overview}

@@

@@

    To our knowledge, this is the first study to investigate the gravitational interactions between all Galactic globular clusters. While we did not  observe significant changes to orbital properties of the clusters, when subject to their mutual interactions -- a result which is  expected given the small sizes and vast distances between the clusters -- we do see clear effects of the presence of their gravitational field in the properties of Pal~5 stream. A particularly clear example of this effect is shown in Fig.~\ref{fig:stream_on_sky} which we selected for its prominent gaps. Two of these gaps are clearly visible in galactic coordinates and become even more apparent when marginalizing over latitude to reconstruct the 1D density profile of the stream, as a function of longitude.

    To make a more quantitative comparison between the \texttt{reference} and \texttt{full} simulations, we adopted to work in the tail coordinate system, in which  the stream is aligned with the cluster's orbit. The definition of this coordinate system is based on the work of \citet{2004AJ....127.2753D} and is shown in Fig.~\ref{fig:TailCoordinates}. Briefly, in this system the x' coordinate represents the position of a particle along the orbit relative to the globular cluster—whether it is ahead (positive x' ) or behind (negative x') the cluster;  the y' coordinate measures the particle's distance within the orbital plane, where positive values indicate that the particle is farther from the Galactic center and negative values indicate that it is closer. Fig.~\ref{fig:decomposition} shows a comparison between one of the 50 realizations of Palomar~5 stream taking into account the gravitational interactions with all other globular clusters in the Galaxy (top panel) and omitting them (bottom panel). This comparison clearly shows the presence of two wide ($\sim$ 100~pc and $\sim$ 1~kpc) gaps in the leading tail and of a more subtle underdensity at x'$\sim 5$~kpc (we refer the reader to Appendix~\ref{sec:gap_detection} for a detailed description of the underdensities and gaps detection method). To find out which globular clusters were responsible of creating these gaps, and at what time the close passages between these clusters and the Palomar~5 streams took place we estimated the gravitational acceleration along the orbit of Pal~5 and represented it in the ($t$, $\tau$) space, where $t$ is the simulation time, from 5~Gyr in the past to the present day, and $\tau$ is indicates how long it will take for Pal~5 to reach, or how  long ago it passed, a given point on its orbit. The use of $\tau$ is advantageous because the growth of the stream is approximately linear in $\tau$ whereas in physical space, streams on eccentric orbits expand and contract depending on the orbital phase. Adopting this time space, and reporting the gravitational acceleration along Pal~5 orbit in this space, the identification of the globular clusters which produced the perturbation, and the time at which it occurred becomes straightforward (we refer the reader to Appendix~\ref{sec:Perturber_Identification} for a in-depth description of the procedure). In this way, we could identify that the clusters responsible for creating gaps in the simulation of Palomar~5 stream reported in Fig.\ref{fig:decomposition} are NGC~2808, NGC~7078 and NGC~104, and that their close passages occurred 200~Myr, \paola{xx and xx Gyr ago}, respectively. To further probe that the three clusters above are responsible of producing the gaps observed in this simulation, we conducted additional experiments by including only the perturbation of each of these three clusters at a time, and neglecting the gravitational perturbations of all the other clusters in the Galaxy. The results of these simulations are reported in the middle panels of Fig.~\ref{fig:decomposition} and clearly shows that NGC~2808, NGC~7078 and NGC~104 are the clusters responsible of creating the underdense regions observed in the leading tails of Palomar~5. It is worth noting that the times at which the passages of these clusters occurred, according to our analysis, are in agreement with the observed width of the corresponding gaps: the encounter with NGC~2808 being very recent (only 200~Myr ago) its induced gap is still very thin, because it takes time for a perturbation to grow into an extended gap, as it is the case for those induced by the passages of  NGC~7078 and NGC~104, which occurred in earlier times. It is also interesting to emphasize that gaps as thin as those generated by the passage of NGC~2808, 200~Myr ago, can be detected by working in the tail coordinate system and by making a differential analysis (\texttt{full} versus \texttt{reference} simulations): they are so thin that they cannot be directly identified in galactic coordinates (see Fig.~\ref{fig:stream_on_sky}).\\

    The analysis presented in Fig.~\ref{fig:decomposition} has been repeated for the whole set of simulations, and is reported in Appendix~\ref{sec:gallery_of_gaps} for completeness. From this analysis we can derive a statistical view of: 1) the number of gaps generated on Pal~5 stream by the whole system of Galactic globular clusters; 2) the clusters which generated these gaps and the time history of these perturbations. From this, we can then quantify: 3) the properties of the perturbers (their masses, sizes and their orbital parameters) as well as 4) the impact geometry of the encounters, which allows us to understand which encounters are more favorable to generate gaps in Pal~5 stream. In the following, we will present results of this analysis, by starting from addressing points 1) and 2).

  \subsection{The history and statistics of gap creations in Palomar~5 stream}\label{sect:history}

    By applying the methodology described above to the whole set of simulations, we can reconstruct the history of close passages of Galactic globular clusters to Palomar~5 stream, in the last 5~Gyr, which -- we recall the reader -- is the time interval investigated in our simulations. The results of this analysis are reported in Fig.~\ref{fig:histogram_impact_time} and in Table~xx \paola{I suggest to report the results of Fig 3 also in a Table, in which we report for each GC which produced a gap, the time of close passage and the total number of gaps it generated.} In total, we report the finding of 73 gaps in our 50 simulations, that is 1.5 gaps/simulation, on average, created by 18 different perturbers. NGC~2808 impacts Palomar 5's stream about 200 Myr ago in 44 out of 50 simulations, creating a small gap at a similar position to the one reported in Fig.~\ref{fig:decomposition}  in each case. Since this interaction occurred less than one orbital period ago, the orbital solutions remain similar, producing consistent results across all simulations, within the uncertainties.  For encounters which took place at earlier times, the clusters responsible of creating the gaps may vary -- that is, gaps are not created at the same time in all simulations and the parameters of a specific encounter between a  globular cluster and Palomar~5 stream may vary significantly from a simulation to another with the turning back of time, so that in some cases a cluster may produce a gap and in some others the perturbation is not strong or close enough to produce any detectable signature in the Palomar~5 stream.  In total, beside NGC~2808, we report the creation of 24 gaps in Pal~5 tails, over the last 3~Gyr generated by \paola{15: check this value, I derived it by eye} different clusters, which corresponds to a rate of 0.8 gaps/100~Myr (note that, including NGC~2808, this rate increases to 29.5 gaps/100~Myr). At even earlier times, the probability of creating gaps in Palomar~5 streams diminishes significantly, but one must take into account that at these times the tails of Pal~5 have not yet had time to develop sufficiently, and the probability of close encounters with other clusters consequently decreases. Overall, across 50 simulations, as previously stated, we observe 73 gaps. The distribution of simulations with a given number of gaps is provided in Table~\ref{table:gap_distribution}.  If we consider NGC~2808 to be an outlier and exclude it from the experiment, we observe an average of 0.6 gaps per simulation. Including all observed gaps results in an average of 1.5 gaps per simulation. Interestingly, most gaps appear in the leading tail, which we consider coincidental \paola{I am still surprised by the high occurrence of gaps in the leading tail, but I amm still not able to come up with an explanation for this. I mean, is it really coincidental? I propose to leave the discussion as it is for the moment. We will add more, if we come up with an other possible explanation}, as stars escape the cluster's L1 and L2 points at similar rates, and the small distance difference between tails' galactocentric distance is negligible. We expect gaps to be equally likely in both tails. However, only 8 of the 73 gaps occur in the trailing tail, with 44 linked to NGC~2808's flyby. After removing these, there are 21 gaps in the leading tail and 8 in the trailing. By the binomial distribution, the probability of observing up to 8 successes in 29 trials, given a 50\% chance of success, is 1.2\%—unlikely, but possible. Additionally, other perturbers impact at consistent times, which may also violate the assumption of independent events, as seen with NGC~2808.

    \begin{table}[h]
      \centering
        \begin{tabular}{|l|c|c|c|c|c|}
          \hline
          Number of Gaps & 0 & 1 & 2 & 3 & 4 \\
          \hline
          Number of Sims. & 4 & 25 & 16 & 4 & 1 \\
          \hline
        \end{tabular}
        \vspace{0.5cm}
      \caption{Occurence of gaps in Palomar~5 streams,  in our simulations. More specifically, the table reports the number of simulations (second row) for which a given number of gaps (first row) is found. }\label{table:gap_distribution}
    \end{table}

    Following  \citet{2012ApJ...748...20C} we can also estimate the rate at which gaps are created per unit stream length. For our simulations, this rate is given by: \begin{equation} \label{eq:gap_creation_rate} \mathcal{R}_{\textrm{Pal 5}} =  \frac{1}{T}\int_{0}^T l^{-1}(t) \sum_i \delta(t-t_i) dt,\end{equation}where $T$ is the total integration time, $l(t)$ is the length of the stream, and $\sum_i \delta(t-t_i)$ sums over the gap occurrences, with $i$ indexing over the number of gaps in a given simulation. Here, $\delta$ represents the Dirac delta function. This expression can be simplified to:\begin{equation}\mathcal{R}_{\textrm{Pal 5}} =  \frac{1}{T} \sum_i \frac{1}{l (t_i)}. \end{equation}

    This computation allows us to analyze the distribution of gap creation rates across all simulations. The results are shown in Fig.~\ref{fig:gapcreationrate}. This is roughly consistent with a simple estimate of the average gap creation rate: with 72\paola{73?} gaps over 5 Gyr of integration time for a stream about 20 kpc in length, the naive estimate is approximately 0.015~km~s$^{-1}$~kpc$^{-2} = 0.015~\rm{Gyr^{-1}kpc^{-1}}$ . This is about double the weighted mean gap creation rate of 0.009~km~s$^{-1}$~kpc$^{-2}$. The naive estimate is higher because it does not account for the growth of the stream over time, unlike Eq.\ref{eq:gap_creation_rate}, where the likelihood of impacting a shorter stream is lower.

    \begin{figure}
      \centering
      \includegraphics[width=\linewidth]{gap_creation_rate.png}
      \caption{The distribution of the number of gaps normalized over the total integration time and unit stream length, as described by Eq.\ref{eq:gap_creation_rate} for the whole set of 50 \texttt{full} simulations. }
      \label{fig:gapcreationrate}
    \end{figure}

  \subsection{Impact geometry and parameters of the perturbers}\label{sect:geometry}

    With the perturbers identified, we perform statistics in the pursuit of understanding what conditions are necessary for a globular cluster  in order to form a gap on Pal~5 stream. We turn to impact theory, which in its simplest form is presented in works such as \citet{2008gady.book.....B}. Consider two particles: one stationary and the other moving past it. The distance between the two particles at their closest approach is known as the impact parameter. To simplify the analysis, the \textit{impulse approximation} is employed, which assumes that the velocity of the perturber remains unchanged during the interaction. This assumption simplifies the computation.

    To understand how the impacted particle is perturbed, one needs to compute its change in momentum, which is determined by integrating the force acting on the particle over the duration of the interaction. A useful approximation for this change in momentum \paola{(per unit mass)} is the force at the closest approach multiplied by an estimate of the interaction time:
    \begin{equation} \label{eq:change_in_momentum} \Delta p \approx \text{Force} \times \text{interaction time} = \frac{GMm}{b^2} \times 2\frac{b}{v} = 2\frac{GM}{bv}, \end{equation}where $M$ is the mass of the perturber, $m$ is the mass of the particle, $b$ is the impact parameter, $v$ is the relative velocity of the perturber with respect to the particle, and $G$ is the gravitational constant. \paola{ATTENTION : in the last formula there is no dependence on $m$, as  for a change in momentum per unit mass (hence my comment few lines before). If you want to keep it as momentum per unit mass, that is a change in velocity, you should also remove $m$ from the other expression (i.e. GMm)}

    From this analysis, we can conclude that a more massive perturber, passing closer to the particle, and moving more slowly, will have a greater impact. It is important to note that the momentum change is inversely proportional to the velocity of the perturber, which contrasts with the intuition from elastic collisions, such as those between billiard balls, where higher velocities result in greater impacts.

    \citet{2015MNRAS.450.1136E} extended this impact theory from one point mass impacting another point mass, to studying how an extended body impacts a stream. This quantifies the change in momentum of a given particle as a function of its distance from the point of greatest impact. In their analysis, the perturber was modeled as a Plummer sphere, like in our simulations. Since the stream is not zero-dimensional and has length, both the parallel and perpendicular components of the pertuber velocity must be considered. Consequently, five parameters determine the change in velocity of a given stream particle: $M$, $r_p$, $b$, $W_\parallel$, and $W_\perp$. As detailed in Appendix~\ref{sec:reconstruction}, we calculated these parameters for all our \texttt{full} simulations, by selecting -- for each of them --  the strongest 5 flybys of a perturber with the Pal~5 stream. Thus, we compute a total of  250 impacts, and flag those that give way to gaps.

    Visual inspection of the five key impact parameters ($M$, $r_p$, $b$, $W_\parallel$, and $W_\perp$) did not reveal a clear distinction between flybys that create gaps and those that do not. Due to the clutter, we omit this graph and instead present... In Fig.\ref{fig:impact_geometry_statistics} we report the result of this analysis, by showing the distribution of these impact variables for fly-bys which give rise to gaps and compare it to the distributions of fly-bys which do not lead to gap formation. Note that in this figure, we show the total relative velocity rather than separating parallel and perpendicular components, because from a visual inspection, we could not find any specific trends when the two velocity components were plotted separately. We also exclude the characteristic cluster radius, as it showed little correlation with the results, likely due to the narrow range of globular cluster radii (see Fig.\ref{fig:mass_size_plane}). This factor might be more significant for dark matter subhalos, where size variation is greater.

    While Fig.~\ref{fig:impact_geometry_statistics} demonstrates that mass, relative velocity, or impact parameter alone (from Eq.~\ref{eq:change_in_momentum}) cannot predict gap formation, one interesting result emerges: no gaps are created from interactions with impact parameters greater than 200~pc. The stream widths are roughly 200~pc, with a radial thickness\paola{what do you mean by radial thickness?} of about 100~pc. This finding is even more evident when examining the $b$-$M$ plane. A series of perturbers at roughly $\sim7 \times 10^5 M_\odot$ highlights NGC~2808's flybys, where all encounters with impact parameters under 200~pc result in gaps, while those beyond this distance do not. In other words, massive globulars with masses greater than $10^6 \rm{M}_\odot$ are not effective in producing gaps if they pass close to Pal~5 stream with an impact parameter $b > 200$~pc. It is interesting to note that for values of the impact parameter below this threshold, even fast encounters ($\delta v > 300$~km/s) can produce gaps. \paola{Attention: there is something I do not understand in Fig 5: the range of $\delta v$ values seems not the same in the first row and in the last row (left panel). Is there anything to clarify/correct?}

    \begin{figure}
      \centering
      \includegraphics[width=1\linewidth]{impact_geometry_statistics_deltaP.png}
      \caption{The distribution of imparted  change in momentum (per unit mass) from a cluster flyby given by Eq.~\ref{eq:change_in_momentum}. The data set includes the top 5 strongest flybys from each simulation. Those that cause gaps are colored while those that do not are left in black. Note that the meaning of the colors is the same as in Fig.~\ref{fig:histogram_impact_time}.}
      \label{fig:deltap}
    \end{figure}

    \begin{figure*}
      \centering
      \includegraphics[width=0.45\linewidth]{E_Lz_perturbers.png}
      \includegraphics[width=0.45\linewidth]{mass_size_plane.png}
      \includegraphics[width=0.45\linewidth]{impact_stats_phase_space.png}
      \caption{\textit{Characterisitcs of gap causing clusters} \textit{Top Left:} Energy-angular momentum space of the globular clusters in the simulations, with 50~$\times$~165 data points representing all sampled initial conditions. Clusters impacting Palomar 5 are shown with colored markers; non-gap-causing clusters are shown in gray. Palomar 5 is marked with a white symbol and black edge. The circular velocity-curve is drawn with a light-pink dashed curve. \textit{Top Right:} Mass-size plane of the globular clusters in the simulations, with uncertainties on the masses indicated with vertical bars. gap-causing clusters are indicated with colored points, whereas Pal~5 is indicated by a white symbol with a black edge. \textit{Bottom Right:} Palomar 5's orbit in the meridional plane. The grayscale represents all 50 stacked orbits, with $\mathcal{P}$ indicating the probability of  Palomar~5's position, normalized to $\mathcal{P}_\textrm{max}=1$. Markers show the perturber's position during each impact. In all panels the colors of the markers and histogram bars are consistent with Fig.~\ref{fig:histogram_impact_time}.\paola{can you add also a panel with Pal~5 orbit and impacts in the (x-y) plane?}}
      \label{fig:mass_size_plane}
    \end{figure*}

  Once all the key impact parameters estimated, we can use Eq~\ref{eq:change_in_momentum} and calculate $\Delta p$, the change in momentum (per unit mass) imparted by a cluster flyby on Pal~5 stream. The result  is shown in Fig.~\ref{fig:deltap}, where the distribution of imparted in change in momentum is shown for all impacts which produce a gap, and compared to those that do not produce one. On average, encounters which lead to gap creations impart a change in momentum on stream particles which is a factor of 10 higher than that of encounters which do not form gaps (but with some overlap in the low-velocity tail). Interestingly, changes in momentum which lead to gap creations extend over a large range in velocities (there is a factor of about 100 between the smallest and largest changes), with NGC~2808 (orange color in the histogram) imparting changes in the velocity of stream particles which redistribute over the whole range of $\Delta p$. \paola{I suggest to use $\Delta p$ instead of $\Delta v$ in the xlabel of Fig.~\ref{fig:deltap}, to avoid confusion with $\delta v$ of Fig.\ref{fig:impact_geometry_statistics}.}\\

  In addition to characterizing the parameters governing cluster encounters with the stream, since we know which clusters have produced gaps on the tail of Pal5, we can also verify their orbital and structural properties.  This is what is shown in Fig~\ref{fig:mass_size_plane}, where we first show the mass and size (i.e. half-mass radius) of the clusters that cause gaps on the tails of Pal5, dividing them from those that do not. As can be seen, no cluster with mass below $10^5 \rm{M}_\odot$ causes gaps on Pal~5 stream, and all perturbers, except xx \paola{add which cluster has a $r_h > 10$pc} have a  half-mass radius below 10~pc. Even more interesting is their distribution in the E-Lz plane, which shows that the clusters that cause gaps are on both direct and retrograde orbits (respectively, negative and positive values of $L_z$), but all confined to an energy interval between $-2$ and $-1 \times10^5~\textrm{km}^2\textrm{s}^{-2}$. This is because only clusters within Palomar 5's orbital space can interact with the stream: clusters with higher orbital energies tend to have larger apocenter than that of Pal~5 and thus spend most of their time away from Palomar 5's orbital volume. Finally,  it is worth commenting on the location of impacts with the stream, whether they occur when Pal~5 is close to its pericenter or not. As it can be seen from Fig.xx (bottom panel) encounters can occur at all orbital phases of Pal~5, when it is close to pericenter, but also very far away from it, at the outskirts of its orbital space. However, when taken all together, the location of gap creating impacts shows a strong negative correlation  with the galactocentric radius $r$, with the number $N$ of encounters favorable to gap formation  going as $N = -2.5r + 50$ (with a Pearson coefficient of -0.86). While there are more clusters near the galactic center, orbits naturally spend more time near their apocenters, and the lower relative velocities in these regions should favor gap creation. However, this result suggests that cluster galactic number density outweighs these factors in determining the number of gaps.

  %Our results are consistent with the findings from \citet{2021MNRAS.505.5978V}, who identified a wide range of orbital solutions for the least bound clusters, including both retrograde and prograde orbits. In our potential model, only three clusters samplings had positive orbital energies.
